Sparse spectral methods for partial differential equations on spherical caps and bands

### 2020-09-02 papers/spherical-caps/spherical-caps.tex
Readthrough

% Spherical Cap commands
\newcommand{\scop}{Q}
\newcommand{\scopnki}{\scop_{n,k,i}}
\newcommand{\scopmjh}{\scop_{m,j,h}}
\newcommand{\scopa}{\scop^{(a)}}
\newcommand{\scopnkia}{\scopnki^{(a)}}
\newcommand{\scopmjha}{\scopmjh^{(a)}}
\newcommand{\bigscop}{{\mathbb{Q}}}
\newcommand{\bigscopa}{\bigscop^{(a)}}
\newcommand{\bigscopN}{\bigscop_{N}}
\newcommand{\bigscopNa}{\bigscopa_{N}}
\newcommand{\bigscopNka}{\bigscopa_{N,k}}
\newcommand{\bigscopt}{\mathbb{\tilde{Q}}}
\newcommand{\bigscopta}{\bigscopt^{(a)}}
\newcommand{\bigscoptn}{\bigscopt_{n}}
\newcommand{\bigscoptna}{\bigscopta_{n}}
\newcommand{\xvec}{\mathbf{x}}
\newcommand{\ch}{Y}
\newcommand{\chki}{\ch_{k,i}}
\newcommand{\chjh}{\ch_{j,h}}
\newcommand{\chkitheta}{\ch_{k,i}(\theta)}
\newcommand{\alphaa}{\alpha^{(a)}}
\newcommand{\betaa}{\beta^{(a)}}
\newcommand{\gammaa}{\gamma^{(a)}}
\newcommand{\Wa}{W^{(a)}}
\newcommand{\bigWNa}{\mathbb{W}_N^{(a)}}
\newcommand{\ppphi}{\frac{\partial}{\partial \phi}}
\newcommand{\ppz}{\frac{\partial}{\partial z}}
\newcommand{\rhoppphi}{\rho \ppphi}
\newcommand{\rhozppphi}{\rho(z) \ppphi}
\newcommand{\pptheta}{\frac{\partial}{\partial \theta}}
\newcommand{\ppthetatwo}{\frac{\partial^2}{\partial \theta^2}}
\newcommand{\bigWNi}{\mathbb{W}_N^{(1)}}
\newcommand{\scopi}{\scop^{(1)}}
\newcommand{\scopnkii}{\scopnki^{(1)}}
\newcommand{\bigWNo}{\mathbb{W}_N^{(0)}}
\newcommand{\scopo}{\scop^{(0)}}
\newcommand{\scopnkio}{\scopnki^{(0)}}
\newcommand{\bigscopi}{\bigscop^{(1)}}
\newcommand{\bigscopNi}{\bigscopi_{N}}
\newcommand{\bigscopNki}{\bigscopi_{N,k}}
\newcommand{\bigscopo}{\bigscop^{(0)}}
\newcommand{\bigscopNo}{\bigscopo_{N}}
\newcommand{\bigscopNko}{\bigscopo_{N,k}}

% Tangent space
\newcommand{\tangentspace}{{\Omega_T}}
\newcommand{\phivec}{\hat{\bm{\phi}}}
\newcommand{\thetavec}{\hat{\bm{\theta}}}
\newcommand{\bigtsop}{\mathbb{T}}
\newcommand{\bigtsopt}{\tilde{\bigtsop}}
\newcommand{\bigtsoptn}{\bigtsopt_n}
\newcommand{\bigtsopN}{\bigtsop_N}
\newcommand{\bigtsopNk}{\bigtsop_{N,k}}
\newcommand{\tsopi}{\bm{\Phi}}
\newcommand{\tsopii}{\bm{\Psi}}
\newcommand{\tsopiN}{\tsopi_N}
\newcommand{\tsopiiN}{\tsopii_N}
\newcommand{\tsopinki}{\tsopi_{n,k,i}}
\newcommand{\tsopiinki}{\tsopii_{n,k,i}}
\newcommand{\tsopinkit}{\tsopi_{n,k,i}^\top}
\newcommand{\tsopiinkit}{\tsopii_{n,k,i}\top}
\newcommand{\tsopit}{\tsopi^\top}
\newcommand{\tsopiit}{\tsopii^\top}
\newcommand{\rvec}{\hat{\bm{r}}}
\newcommand{\jacobimattangent}{{}^T\!J}
\newcommand{\jacobimattangentx}{\jacobimattangent_x}
\newcommand{\jacobimattangenty}{\jacobimattangent_y}
\newcommand{\jacobimattangentz}{\jacobimattangent_z}
\newcommand{\alphao}{\alpha^{(0)}}
\newcommand{\alphaonkj}{\alphao_{n,k,j}}
\newcommand{\betao}{\beta^{(0)}}
\newcommand{\betaonkj}{\betao_{n,k,j}}
\newcommand{\gammao}{\gamma^{(0)}}
\newcommand{\gammaonkj}{\gammao_{n,k,j}}
\newcommand{\uvec}{\bm{u}}
\newcommand{\hvec}{\bm{h}}
\newcommand{\uvecc}{\uvec^c}
\newcommand{\hvecc}{\hvec^c}
\newcommand{\uvecn}{\uvec_n}
\newcommand{\hvecn}{\hvec_n}
\newcommand{\uvecnpi}{\uvec_{n+1}}
\newcommand{\hvecnpi}{\hvec_{n+1}}
\newcommand{\uvecnc}{\uvecc_n}
\newcommand{\hvecnc}{\hvecc_n}
\newcommand{\uvecnpic}{\uvecnpi^c}
\newcommand{\hvecnpic}{\hvecnpi^c}
\newcommand{\href}{\mathcal{H}}
\newcommand{\divergence}{\nabla \cdot}
\newcommand{\grad}{\nabla}

@@

\title{Sparse spectral and $p$-finite element methods for partial differential equations on spherical caps}
\author{Ben Snowball, Sheehan Olver\thanks{Department of Mathematics, Imperial College, London, England}}
%\date{}							% Activate to display a given date or no date

@@

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend the methodology on disk-slices developed in \cite{snowball2019sparse} to a hierarchy of non-classical orthogonal polynomials on spherical caps. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on spherical subdomains such as spherical bands and spherical caps. More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: ||(x,y)|| := \sqrt{x^2 + y^2} = \gamma \rho(z)\}
\end{align*}
and consider the case where $\alpha \in (0,1)$ and where
\begin{align*}
	 \rho(z) &:= \sqrt{1-z^2} \\
	 \beta &:= 1 \\
	 \gamma &:= 1.
\end{align*}
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < 1, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
Of course, there is a simple extension to a spherical band by taking $\beta \in (\alpha,1)$. However for simplicity we focus on the spherical cap in this paper.

As in the work in two dimensions on the show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that are defined in such a way that mirrors the approach on quadratic curves \cite{olver2020orthogonal}. On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k - \tilde k_i} \: y^{\tilde k_i} \: z^{n-k} + (\hbox{lower order terms})
\end{align*} \bstodoinline{How can I write down what the ordering is? Is this necessary?}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - \tilde j_i} \: y^{\tilde j_i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$. In this work, as for the disk-slice case in 2D \cite{snowball2019sparse}, we use a more general integration-by-parts argument to deduce the sparsity structure. The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations \bstodoinline{Presumably this is correct to say?}. Furthermore, we can use this framework to derive sparse $p$-finite element methods, which we introduce as an Appendix here. The methods are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

@@

\section{Orthogonal polynomials on the disk-slice and the trapezium}\label{Section:OPs}

In this section we outline the construction and some basic properties of $\scopnkia(x,y)$.

@@

We can construct the 3D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the interval \([\alpha,\beta]\), and from Chebyshev polynomials.

\begin{definition}
Define the unit circle $\omega := \{ \mathbf{x} = (x,y) \in \R^2 \: | \: x^2 + y^2 = 1\}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos(\theta)$, $y = \sin(\theta)$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\begin{align}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \nonumber \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \ &:= T_k(x) = \cos(k \theta) =: \ch_{k,0}(\theta), \quad k = 1,2,3,\dots\label{eqn:chebyshev} \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin(k \theta) =: \ch_{k,1}(\theta), \quad k = 1,2,3,\dots \nonumber
\end{align}
where $\ch_{0} := \frac{\sqrt{2}}{2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_{||\xvec|| = 1} p(\xvec) \: q(\xvec) \: \D \sigma(\xvec) = \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta. \label{eqn:ipcircularharmonics}
\end{align}
\end{definition}
Note that we have defined $\ch_0$ so as to ensure orthonormality.

\begin{proposition}\label{prop:construction}
Let $w : (\alpha,\beta) \: \to \R$ be a weight function. $\forall$, $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) \equiv \scopnki(\xvec,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr(\frac{\xvec}{\rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: w(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: w(z) \: \D z
\end{align*}
on $\Omega$. 
\end{proposition}

@@

By construction, $\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z,
\end{align}
with
\begin{align}
	\norm{\scopnkia}_{\scopa}^2 &:= \ip< \scopnkia, \: \scopnkia >_{\scopa} = \pi \: \normgenjac^{(a, 2k)} \label{eqn:normscop}
\end{align}
We note that the weight $\genjacw^{(a,b)}(z)$ has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}, where a method for obtaining recurrence coefficients and evaluating integrals was established (the weight is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).

For clarity we state here that throughout we will regularly use coordinate transformations, mainly in any integrals over the surface $\Omega$, between the cartesian $(x,y,z)$ and the spherical coordinates $(\phi, \theta)$ -- that is, for any point $(x,y,z) \in \Omega$, let $(\phi, \theta)$ denote the angles such that:
\begin{align*}
	z &= \cos(\phi) \\
	x &= \sin(\phi) \cos(\theta) = \rho(z) \cos(\theta) \\
	y &= \sin(\phi) \sin(\theta) = \rho(z) \sin(\theta).
\end{align*}

@@

In this section, we introduce new notation for a different ordering of the OP vector, in order to demonstrate how we can obtain a recurrence for the OPs in matrix-vector form, and derive the associated \enquote{Clenshaw matrices}. While in Section \ref{subsection:jacobimats} the ordering used is convenient for the application of Jacobi and differential operators to the vector of coefficients of a given function's expansion, the ordering we will use in this section is convenient for building the OPs and more importantly obtaining the \enquote{Clenshaw matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings.

@@

\begin{proof}[Proof of Theorem \ref{theorem:sparsityofdifferentialoperators}]
First, note that:
\begin{align}
	\genjacw^{(a,b) \: \prime}(z) &= a \: \genjacw^{(a-1,b)}(z) + c \: \rho(z) \: \rho'(z) \:\genjacw^{(a,b-2)}(z), \label{eqn:derivativeofweightgenjac} \\
	\rho(z) \: \rho'(z) &= -z \label{eqn:rhoderivative} \\
	{\partial \over \partial \phi} &= -\rho \: {\partial \over \partial z} \label{eqn:partialphitoz} 
\end{align}

For the operator $D_\theta$ for partial differentiation by $\theta$, we simply have that
\begin{align*}
	{\partial \over \partial \theta}\scopnkia(x,y,z) &= \genjacnmk^{(a,2k)}(z) \: \rho(z)^{k} \: {\partial \over \partial \theta} \chki(\theta) \\
	&= 
	\begin{cases}
		(-1)^{i+1} \: k \: \scopa_{n,k,|i-1|}(x,y,z) & k > 0 \\
		0 & k = 0
	\end{cases}
\end{align*}

>>>>

% Spherical Cap commands
\newcommand{\scop}{Q}
\newcommand{\scopnki}{\scop_{n,k,i}}
\newcommand{\scopmjh}{\scop_{m,j,h}}
\newcommand{\scopa}{\scop^{(a)}}
\newcommand{\scopnkia}{\scopnki^{(a)}}
\newcommand{\scopmjha}{\scopmjh^{(a)}}
\newcommand{\bigscop}{{\mathbb{Q}}}
\newcommand{\bigscopa}{\bigscop^{(a)}}
\newcommand{\bigscopN}{\bigscop_{N}}
\newcommand{\bigscopNa}{\bigscopa_{N}}
\newcommand{\bigscopNka}{\bigscopa_{N,k}}
\newcommand{\bigscopt}{\mathbb{\tilde{Q}}}
\newcommand{\bigscopta}{\bigscopt^{(a)}}
\newcommand{\bigscoptn}{\bigscopt_{n}}
\newcommand{\bigscoptna}{\bigscopta_{n}}
\newcommand{\xvec}{\mathbf{x}}
\newcommand{\ch}{Y}
\newcommand{\chki}{\ch_{k,i}}
\newcommand{\chjh}{\ch_{j,h}}
\newcommand{\chkitheta}{\ch_{k,i}(\theta)}
\newcommand{\alphaa}{\alpha^{(a)}}
\newcommand{\betaa}{\beta^{(a)}}
\newcommand{\gammaa}{\gamma^{(a)}}
\newcommand{\Wa}{W^{(a)}}
\newcommand{\bigWNa}{\mathbb{W}_N^{(a)}}
\newcommand{\ppphi}{\frac{\partial}{\partial \varphi}}
\newcommand{\ppz}{\frac{\partial}{\partial z}}
\newcommand{\rhoppphi}{\rho \ppphi}
\newcommand{\rhozppphi}{\rho(z) \ppphi}
\newcommand{\pptheta}{\frac{\partial}{\partial \theta}}
\newcommand{\ppthetatwo}{\frac{\partial^2}{\partial \theta^2}}
\newcommand{\bigWNi}{\mathbb{W}_N^{(1)}}
\newcommand{\scopi}{\scop^{(1)}}
\newcommand{\scopnkii}{\scopnki^{(1)}}
\newcommand{\bigWNo}{\mathbb{W}_N^{(0)}}
\newcommand{\scopo}{\scop^{(0)}}
\newcommand{\scopnkio}{\scopnki^{(0)}}
\newcommand{\bigscopi}{\bigscop^{(1)}}
\newcommand{\bigscopNi}{\bigscopi_{N}}
\newcommand{\bigscopNki}{\bigscopi_{N,k}}
\newcommand{\bigscopo}{\bigscop^{(0)}}
\newcommand{\bigscopNo}{\bigscopo_{N}}
\newcommand{\bigscopNko}{\bigscopo_{N,k}}

% Tangent space
\newcommand{\tangentspace}{{\Omega_T}}
\newcommand{\phivec}{\hat{\bm{\varphi}}}
\newcommand{\thetavec}{\hat{\bm{\theta}}}
\newcommand{\bigtsop}{\mathbb{T}}
\newcommand{\bigtsopt}{\tilde{\bigtsop}}
\newcommand{\bigtsoptn}{\bigtsopt_n}
\newcommand{\bigtsopN}{\bigtsop_N}
\newcommand{\bigtsopNk}{\bigtsop_{N,k}}
\newcommand{\tsopi}{\bm{\Phi}}
\newcommand{\tsopii}{\bm{\Psi}}
\newcommand{\tsopiN}{\tsopi_N}
\newcommand{\tsopiiN}{\tsopii_N}
\newcommand{\tsopinki}{\tsopi_{n,k,i}}
\newcommand{\tsopiinki}{\tsopii_{n,k,i}}
\newcommand{\tsopinkit}{\tsopi_{n,k,i}^\top}
\newcommand{\tsopiinkit}{\tsopii_{n,k,i}\top}
\newcommand{\tsopit}{\tsopi^\top}
\newcommand{\tsopiit}{\tsopii^\top}
\newcommand{\rvec}{\hat{\bm{r}}}
\newcommand{\jacobimattangent}{{}^T\!J}
\newcommand{\jacobimattangentx}{\jacobimattangent_x}
\newcommand{\jacobimattangenty}{\jacobimattangent_y}
\newcommand{\jacobimattangentz}{\jacobimattangent_z}
\newcommand{\alphao}{\alpha^{(0)}}
\newcommand{\alphaonkj}{\alphao_{n,k,j}}
\newcommand{\betao}{\beta^{(0)}}
\newcommand{\betaonkj}{\betao_{n,k,j}}
\newcommand{\gammao}{\gamma^{(0)}}
\newcommand{\gammaonkj}{\gammao_{n,k,j}}
\newcommand{\uvec}{\bm{u}}
\newcommand{\hvec}{\bm{h}}
\newcommand{\uvecc}{\uvec^c}
\newcommand{\hvecc}{\hvec^c}
\newcommand{\uvecn}{\uvec_n}
\newcommand{\hvecn}{\hvec_n}
\newcommand{\uvecnpi}{\uvec_{n+1}}
\newcommand{\hvecnpi}{\hvec_{n+1}}
\newcommand{\uvecnc}{\uvecc_n}
\newcommand{\hvecnc}{\hvecc_n}
\newcommand{\uvecnpic}{\uvecnpi^c}
\newcommand{\hvecnpic}{\hvecnpi^c}
\newcommand{\href}{\mathcal{H}}
\newcommand{\divergence}{\nabla \cdot}
\newcommand{\grad}{\nabla}

\def\DeltaS{\Delta_{\rm S}}

@@

\title{Sparse spectral methods for partial differential equations on spherical caps and bands}
\author{Ben Snowball, Sheehan Olver\thanks{Department of Mathematics, Imperial College, London, England}}
%\date{}							% Activate to display a given date or no date

@@

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps and bands. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and Biharmonic operators, showing spectral convergence. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on certain subsets of the sphere: spherical bands and spherical caps. More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: ||(x,y)|| := \sqrt{x^2 + y^2} =  \rho(z)\}
\end{align*}
and consider the case where \sotodo{Why is $\alpha$ restricted to be positive?} $\alpha \in (0,1)$ and where
$ \rho(z) := \sqrt{1-z^2}$,  $\beta := 1$ \sotodo{Don't we need $\beta \neq 1$ to get a band?}.
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < 1, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
Of course, there is a simple extension to a spherical band by taking $\beta \in (\alpha,1)$. However for simplicity we focus on the spherical cap in this paper.

We advocate using a basis that is polynomial in cartesian coordinates, that is, polynomial in $x$, $y$, and $z$, and orthogonal with respect to a prescribed weight: that is, multivariate orthogonal polynomials, whose construction was considered in \cite{olver2020orthogonal}. Equivalently, we can think of these as polynomials modulo the vanishing ideal $\{ x^2 + y^2 + z^2 = 1 \}$, or simply as a linear recombination of spherical harmonics that are orthogonalised on a subset of the sphere.    This is in contrast to more standard approaches based on mapping the geometry to a simpler one (e.g., a rectangle or disk) and using orthogonal polynomials in the mapped coordinates (e.g., a basis that is polynomial in the spherical coordinates $\varphi$ and $\theta$). The benefit of the new approach is that we do not need to resolve Jacobians, and thereby we can achieve sparse discretisations for partial differential operators, including those with polynomial variable coefficients.   On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is, \sotodo{I don't understand this ordering. Shouldn't it either be $x^k y^j z^{n-k-j}$ or $x^k y^{j-k} z^{n-j}$?  PS I prefer $j$ to $i$ }
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k - \tilde k_i} \: y^{\tilde k_i} \: z^{n-k} + (\hbox{lower order terms})
\end{align*} \bstodoinline{How can I write down what the ordering is? Is this necessary?}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - \tilde j_i} \: y^{\tilde j_i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

We consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): in spherical coordinates 
\begin{align*}
	z &= \cos\varphi, \\
	x &= \sin\varphi \cos\theta = \rho(z) \cos\theta, \\
	y &= \sin\varphi \sin\theta = \rho(z) \sin\theta.
\end{align*}
we have \sotodo{Double check}
\begin{align*}
\DeltaS &= {1 \over \sin \varphi} {\partial \over \partial \varphi} \sin \varphi {\partial \over \partial \varphi} + {1 \over \sin^2 \varphi} {\partial^2 \over \partial \theta^2}  = {1 \over \rho} {\partial \over \partial \varphi} \rho {\partial \over \partial \varphi} + {1 \over \rho^2} {\partial^2 \over \partial \theta^2} \\
\end{align*}
We do so by considering the component operators $\rho {\partial \over \partial \varphi}$ and ${\partial \over \partial \theta}$ applied to OPs with a specific choices of weight so that their discretisation is sparse, see Theorem~\ref{theorem:sparsityofdifferentialoperators}.  Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$, a la the ultraspherical spectral method for intervals \cite{olver2013fast}, triangles \cite{olver2019triangle} and disk-slices and trapeziums \cite{snowball2019sparse}, and the related work on sparse discretisations on disks \cite{vasil2016tensor} and spheres \cite{vasil2019tensor,lecoanet2019tensor}.  As in the  disk-slice case in 2D \cite{snowball2019sparse}, we use an integration-by-parts argument to deduce the sparsity structure.

 The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical\sotodo{Add citation}) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations. Furthermore, we can use this framework to derive sparse $p$-finite element methods, which we introduce as an Appendix here. The methods are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

Note that we consider partial differential operators that are not necessarily rotational invariant:  for example, one can use these techniques for  Schr\"odinger operators $\DeltaS + V(x,y,z)$ where $V$ is first approximated by a polynomial. A nice feature though is that if the partial differential operator is invariant with respect to rotation around the $z$ axes (e.g., a Schr\"odinger operator with potential $V(z)$) the discretisation decouples, and can be reordered as a block-diagonal matrix. This improves the complexity further to an optimal $O(p^2$). \sotodo{Adding a complexity plot showing this would be great}

@@

\section{Orthogonal polynomials on spherical caps}\label{Section:OPs}

In this section we outline the construction and some basic properties of $\scopnkia(x,y,z)$.

@@

We can construct the 3D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the interval \([\alpha,\beta]\), and from Chebyshev polynomials. We do so in terms of Fourier series, which, following \cite{olver2020orthogonal}, we write here as orthogonal polynomials in $x$ and $y$:

\begin{definition}
Define the unit circle $\omega := \{ \mathbf{x} = (x,y) \in \R^2 \: | \: x^2 + y^2 = 1\}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\begin{align}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \nonumber \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \ &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta), \quad k = 1,2,3,\dots\label{eqn:chebyshev} \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta), \quad k = 1,2,3,\dots \nonumber
\end{align}
where $\ch_{0} := \frac{\sqrt{2}}{2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product\sotodo{The $\sigma$ notation needs explanation}
\begin{align}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_{||\xvec|| = 1} p(\xvec) \: q(\xvec) \: \D \sigma(\xvec) = \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta. \label{eqn:ipcircularharmonics}
\end{align}
\end{definition}
Note that we have defined $\ch_0$ so as to ensure orthonormality.

\sotodoinline{Isn't the following from \cite{olver2020orthogonal}? We then don't need to repeat the proof}

\begin{proposition}\label{prop:construction}
Let $w : (\alpha,\beta) \: \to \R$ be a weight function. $\forall$, $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) \equiv \scopnki(\xvec,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr(\frac{\xvec}{\rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: w(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos\theta, \rho(z) \sin\theta, z) \: q(\rho(z) \cos\theta, \rho(z) \sin\theta, z) \: \D \theta \: w(z) \: \D z
\end{align*}
on $\Omega$. 
\end{proposition}

@@

By construction, $\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos\theta, \rho(z) \sin\theta, z) \: q(\rho(z) \cos\theta, \rho(z) \sin\theta, z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z,
\end{align}
with
\begin{align}
	\norm{\scopnkia}_{\scopa}^2 &:= \ip< \scopnkia, \: \scopnkia >_{\scopa} = \pi \: \normgenjac^{(a, 2k)} \label{eqn:normscop}
\end{align}
We note that the weight $\genjacw^{(a,b)}(z)$ has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}, where a method for obtaining recurrence coefficients and evaluating integrals was established (the weight is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).

@@

In this section, we introduce new notation for a different ordering of the OP vector, in order to demonstrate how we can obtain a recurrence for the OPs in matrix-vector form, and derive the associated\sotodo{Explain what a Clenshaw matrix is} \enquote{Clenshaw matrices}. While in Section \ref{subsection:jacobimats} the ordering used is convenient for the application of Jacobi and differential operators to the vector of coefficients of a given function's expansion, the ordering we will use in this section is convenient for building the OPs and more importantly obtaining the \enquote{Clenshaw matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings.

@@

\begin{proof}[Proof of Theorem \ref{theorem:sparsityofdifferentialoperators}]
First, note that \sotodo{Here and elsewhere: ${\partial \over \partial z}$ is not well-defined as we are on the sphere. Though perhaps ${\rm d} \over {\rm d} z$ works as we are differentiating the formula with respect to $z$?}:
\begin{align}
	\genjacw^{(a,b) \: \prime}(z) &= a \: \genjacw^{(a-1,b)}(z) + c \: \rho(z) \: \rho'(z) \:\genjacw^{(a,b-2)}(z), \label{eqn:derivativeofweightgenjac} \\
	\rho(z) \: \rho'(z) &= -z \label{eqn:rhoderivative} \\
	{\partial \over \partial \varphi} &= -\rho \: {\partial \over \partial z} \label{eqn:partialphitoz} 
\end{align}

For the operator $D_\theta$ for partial differentiation by $\theta$, we simply have that \sotodo{Would it be better as ${\rm d} \over {\rm d} \theta$?}
\begin{align*}
	{\partial \over \partial \theta}\scopnkia(x,y,z) &= \genjacnmk^{(a,2k)}(z) \: \rho(z)^{k} \: {\partial \over \partial \theta} \chki(\theta) \\
	&= 
	\begin{cases}
		(-1)^{i+1} \: k \: \scopa_{n,k,|i-1|}(x,y,z) & k > 0 \\
		0 & k = 0
	\end{cases}
\end{align*}
